---
title: "Braya-sø"
author: Gavin Simpson
date: February 2017
fontsize: 11pt
geometry: margin=1in
output:
  pdf_document:
    highlight: tango
    fig_width: 6
    fig_height: 3.5
    fig_crop: false
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
knitr::opts_chunk$set(comment=NA, fig.align = "center", out.width = "0.7\\linewidth",
                      echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

## Introduction

In this example, you'll work with the Braya-sø record that I used in the seminar yesterday. First load some packages

```{r packages, echo = TRUE, results = "hide", cache=FALSE}
library("mgcv")
library("ggplot2")
theme_set(theme_bw())
```

```{r braya-so-data, echo=FALSE}
braya <- read.table("../00-data-sets/DAndrea.2011.Lake Braya So.txt", skip = 84)
```

Next load the data

```{r braya-so-data-2, echo=TRUE, eval=FALSE}
braya <- read.tabl("DAndrea.2011.Lake Braya So.txt", skip = 84)
```

We need to do a little data manipulation to fix up the variable names and to add a `sampleInterval` variable. The we prepare a plot of the data using **ggplot2**

```{r braya-so-plot, out.width = "0.9\\linewidth"}
names(braya) <- c("Depth", "DepthUpper", "DepthLower", "Year", "YearYoung", "YearOld", "UK37")
braya <- transform(braya, sampleInterval = YearYoung - YearOld)

## Generate a plot of the data
ylabel <- expression(italic(U)[37]^{italic(k)})
ggplot(braya, aes(x = Year, y = UK37)) +
    geom_line(colour = "grey") +
    geom_point() +
    labs(y = ylabel)
```

The standard GAM we might fit would be a thin-plate spline model. We pass in `sampleInterval` to the `weights` argument. We also need to set the basis quite large in order to capture the true model.

```{r fit-gam-braya-so, out.width = "0.9\\linewidth"}
## Model 6: TPRS, weights as sampleInterval, k = needs to be higher
mod6 <- gam(UK37 ~ s(Year, k = 40, bs = "tp"), data = braya, method = "REML",
            weights = sampleInterval)
summary(mod6)
```

Now we have the fitted model, we can draw the fitted trend using techniques we've seen elsewhere on the course

```{r plot-gam-braya-so}
N <- 500
newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
newYear <- cbind(newYear, data.frame(predict(mod6, newYear, se.fit = TRUE,
                                             unconditional = TRUE)))
newYear <- transform(newYear, upper = fit + (2 * se.fit), lower = fit - (2 * se.fit))

## Draw the fitted spline on the data
plt.fit <- ggplot(braya, aes(x = Year, y = UK37)) +
    geom_ribbon(data = newYear, aes(ymin = lower, ymax = upper, x = Year),
                inherit.aes = FALSE, alpha = 0.25, fill = "steelblue3") +
    geom_line(data = newYear, aes(y = fit, x = Year), size = 1, colour = "steelblue3") +
    geom_point() +
    labs(y = ylabel)
plt.fit
```

The thin-plate spline model overfits the record in some parts. We might try an adaptive spline instead of the thin-plate spline. To change he spline type, we us `bs = "ad"` for adaptive spline

```{r fit-ad-gam-braya-so}
## Model 3: Adaptive spline, weights as sampleInterval
mod3 <- gam(UK37 ~ s(Year, k = 30, bs = "ad"), data = braya,
            method = "REML", weights = sampleInterval)
summary(mod3)
```

We can now plot this model to look at the fitted trend and compare how it fits relative to the thin-plate one

```{r plot-ad-gam-braya-so}
N <- 500
newYear <- with(braya, data.frame(Year = seq(min(Year), max(Year), length.out = N)))
newYear <- cbind(newYear, data.frame(predict(mod3, newYear, se.fit = TRUE,
                                             unconditional = TRUE)))
newYear <- transform(newYear, upper = fit + (2 * se.fit), lower = fit - (2 * se.fit))

## Draw the fitted spline on the data
plt.ad <- ggplot(braya, aes(x = Year, y = UK37)) +
    geom_ribbon(data = newYear, aes(ymin = lower, ymax = upper, x = Year),
                inherit.aes = FALSE, alpha = 0.25, fill = "steelblue3") +
    geom_line(data = newYear, aes(y = fit, x = Year), size = 1, colour = "steelblue3") +
    geom_point() +
    labs(y = ylabel)
plt.ad
```
