---
title: "Small water stable isotopes"
author: Gavin Simpson
date: February 2017
fontsize: 11pt
geometry: margin=1in
output:
  pdf_document:
    highlight: tango
    fig_width: 6
    fig_height: 3.5
    fig_crop: false
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
knitr::opts_chunk$set(comment=NA, fig.align = "center", out.width = "0.7\\linewidth",
                      echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

## Introduction

In this example, you'll work with a $\delta^{15}N$ isotope record from Small Water, a small corrie lake in the UK Lake District. Begin by loading some packages

```{r packages, echo = TRUE, results = "hide", cache=FALSE}
library("mgcv")
library("ggplot2")
theme_set(theme_bw())
```

```{r small-data, echo=FALSE}
small <- readRDS("../00-data-sets/small-water-isotope-data.rds")
head(small)
```

Next load the data

```{r braya-so-data-2, echo=TRUE, eval=FALSE}
small <- readRDS("small-water-isotope-data.rds")
head(small)
```

Next we prepare a plot of the data using **ggplot2**

```{r small-plot, out.width = "0.9\\linewidth"}
ggplot(small, aes(x = Year, y = d15N)) +
    geom_point(colour = "black") +
    theme_bw() +
    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
```

To fit a GAM with an autocorrelation structure we need the `gamm()` function, with an extra "m"

```{r small-model-fit}
## model it using gamm()
mod <- gamm(d15N ~ s(Year, k = 15), data = small,
            correlation = corCAR1(form = ~ Year), method = "REML")
```

The `gamm()` function fits using a mixed model and as such as two different sides, a GAM side and a mixed model side. We need to look at each side to get out relevant information. the `intervals()` function extracts a confidence interval and estimate for the $\phi$ parameter which is the measure of autocorrelation

```{r small-summaries}
summary(mod$gam)
intervals(mod$lme, which = "var-cov")
plot(mod$gam)
```

What does this correlation function look like? Here we pull out details of the structure and draw the corresponding correlation function

```{r cor-struct}
## plot CAR(1) process
maxS <- with(small, diff(range(Year))) ## too large, truncate to 50
S <- seq(0, 50, length = 100)
phi <- intervals(mod$lme, which = "var-cov")$corStruct
car1 <- setNames(as.data.frame(t(outer(phi, S, FUN = `^`)[1, , ])),
                 c("Lower","Correlation","Upper"))
car1 <- transform(car1, S = S)

car1Plt <- ggplot(car1, aes(x = S, y = Correlation)) +
    geom_ribbon(aes(ymax = Upper, ymin = Lower),
                fill = "grey", alpha = 0.3) +
    geom_line() +
    ylab(expression(italic(h) * (list(Delta[t], phi))))
car1Plt
```

The fitted model and data can be plotted using the code below

```{r model-predict-plot}
## Predict from model
newYear <- with(small, data.frame(Year = seq(min(Year), max(Year), length.out = 100)))
newYear <- cbind(newYear, data.frame(predict(mod$gam, newYear, se.fit = TRUE)))

## Draw the fitted spline on the data
plt.fit <- ggplot(small, aes(x = Year, y = d15N)) +
    geom_line(data = newYear, aes(y = fit, x = Year), colour = "black") +
    geom_point(colour = "black") +
    theme_bw() +
    ylab(expression(delta^{15}*N ~(Bulk~Organic~Matter)))
plt.fit
```
