---
title: "Basic Ordination"
author: Gavin L. Simpson
date: U Adelaide 2017 • Feb 13--17 2017
fontsize: 10pt
classoption: "compress, aspectratio=169"
bibliography: vegan-refs.bib
output:
  beamer_presentation:
    theme: metropolis
    keep_tex: true
    highlight: tango
    slide_level: 2
    template: ~/work/teaching/adelaide/feb-2017/slide-template-feb-2017.tex
    fig_width: 6
    fig_height: 3.5
    fig_crop: false
    pandoc_args: [
      "--latex-engine=xelatex"
    ]
---

```{r setup-options, echo = FALSE, results = "hide", message = FALSE}
knitr::opts_chunk$set(comment=NA, fig.align = "center",
                      out.width = "0.7\\linewidth",
                      echo = TRUE, message = TRUE, warning = TRUE,
                      cache = TRUE)
knitr::knit_hooks$set(crop.plot = knitr::hook_pdfcrop)
```

```{r packages, echo = FALSE, results = "hide", message = FALSE}
library("vegan")
data(varespec)
data(varechem)
```

# Introduction

## Introduction



# Unconstrained Ordination

## Unconstrained Ordination


## Links

I have several **vegan**-related posts on my blog. For a list of posts see [http://www.fromthebottomoftheheap.net/blog/]()

## Re-use

Copyright \textcopyright (2015--17) Gavin L. Simpson *Some Rights Reserved*

Unless indicated otherwise, this slide deck is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).

\begin{center}
  \ccby
\end{center}

## References

